---
title: "Simple_stan_inference"
author: "Jennifer Blanc"
date: "11/8/2019"
output: html_document
---

```{r setup, include=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(rstan)
options(mc.cores = parallel::detectCores())
rstan_options(auto_write = TRUE)
set.seed(1)
```

## Introduction  

For this project, we are interested in infering the degree of mutational bias across traits. In order to calculate the mutational bias, we need to infer specific parameters from the data. First consider the polygenic mutation-selection balance model. In this model, the selection coefficient for each individual allele is determined by the approximation below:  

$$s \approx \beta \phi(T) S$$  

Here $\beta$ is the true effect size of the allele, $\phi(T)$ is the height of the liability threshold, and S is the fitness cost of the disease. We can think of $\beta * \phi(T)$ as an approximation of the portion of the population pushed over the liability threshold when the protective allele is swapped to the risk allele. $\phi(T) S$ is a compound parameter that is related to the prevelance of the disease. The larger the $\phi(T)$ the higher the prevalance of disease (the threshold is closer to the center), the smaller the fitness cost must be. Here we will call this compound parameter $W$:  

$$W = \phi(T)S$$  

In this analysis we are interested in infering from simulated data. We will simulate data under a specfic value of W and then use bayesian inference (using STAN) to infer back the value of W, the data were generated under.   

## Infer W using RAF   

To start with we are going to make a (very) simplifying assumption that we know all the causal loci and that they all have an effect size of one ($\beta$ = 1). The distribution of risk alleles in the population for a given effect size $(\beta)$, population size (N), mutation rate $(\mu)$, and W is given below:  

$$\psi(x|\beta, N,W, \mu) = C * e^{-4 N\beta W x} (x(1-x))^{4N\mu -1}$$

The first thing we need to do is simulate data (set of risk allele frequencies) from the above distribution under a specfic value of W. 

Here are helper functions that will allow us to draw allele frequencies from $psi$. We first need a function for the distribution without the constant of integration. For the set of parameters, we will calculate C by integrating this function and then using that C as an argument for the target $\psi$ function we want to draw samples from.  
```{r}
# Function to integrate 
psi_no_c <- function(x, N, B, W, mu) {
  p1 <- exp(-4 * N * B * W * x)
  p2 <- (x *(1-x))^((4*N*mu)-1)
  return(p1 * p2)
}

# This is the target distribution from which we want to sample
psi_target <- function(x, N, B, W, mu, C) {
  p1 <- exp(-4 * N * B * W * x)
  p2 <- (x *(1-x))^((4*N*mu)-1)
  return(p1 * p2 * C)
}
```

This function will simulate a set of Risk Allele frequencies acording to $\psi$ using inverse transform sampling and discritizing the distribution.  
```{r}
simulate_data <- function(N, B, W, m) {
  # This function will sample a set of risk allele frequencies 
  
  # Calculate normalizing constant 
  C <- 1/ integrate(psi_no_c,  (0.5/(2*N)),1 - (0.5/(2*N)) , N=N, B=B, W=W, mu=mu)$value
  
  # Discretize the target CDF 
  i_values <- seq(0,1-(1/(2*N)), 1/(2*N))
  pis <- rep(0, length(i_values))

  # Integrate up to i + (0.5/2N) for all i 
  for (j in 1:length(i_values)) {
    pis[j] <- integrate(psi_target, (0.5/(2*N)), (i_values[j]+ (0.5/(2*N))), N=N, B=B, W=W, mu=mu, C=C)$value
  }
  
  # Draw from uniform
  U <- runif(m)

  # Find smallest i such that pi < U is true
  UT <- rep(0, length(U))
  for (j in 1:length(U)) {
    UT[j] <- which(U[j] <= pis)[1] - 1
  }
  
  # Calculate set of RAF 
  x_freqs <- UT / (2*N)
  
  return(x_freqs)
}
```

Now we are ready to simulate data. We will set the constant parameters we are not interested in inferring and then set W and simulate the data   
```{r}
# Constant parameters 
N <- 100000
mu <- 10^-9 
B <- 1
m <- 500 # Number of causal snps 

# Value of W for simuation 
W <- 1e-5

# Simulate data 
x_freqs <- simulate_data(N, B, W, m)

# Look at distribution of data
hist(x_freqs)
```


Now we have data are are ready to do inference. Let's look again at the distrubtion of risk alleles as stated above:  

$$\psi(x|\beta, N,W, \mu) = C * e^{-4 N\beta W x} (x(1-x))^{4N\mu -1}$$

Using Bayes rule we can re-write this as:  

$$P(W|\beta, N, \mu, x) = \frac{P(x| \beta, N, \mu, W) P(W)}{P(x|\beta, N, \mu)} \propto P(x| \beta, N, \mu, W) P(W)$$  

Now we have a posterior distribution on W that we can sample from using the likelihood ($\psi(x|\beta, N,W, \mu)$) and the prior, which we for now assume to be 1. Below is the STAN program I wrote to sample from this posterior    

**Confusion:** I'm not sure how to deal with the constant of integration in $psi$. To me it seems like in order to evaluate the likelihood we need to recalculate C for each possible step in the MCMC but I'm not sure how to do this in stan.   

```{stan, output.var="simple_RAF"}
functions {
  real my_raf_lpdf(real[] x, real W, int N, real mu, real B){ // PSI density 
    vector[num_elements(x)] prob;
    real lprob;
    for (i in 1:num_elements(x)) {
      prob[i] = ((x[i] * (1- x[i]))^((4*N*mu) - 1)) * exp(-4 * N * B * W * x[i]) ;
    }
    lprob = sum(log(prob));
    return lprob;
  }
}
data {
  int<lower=0> m;         // Number snps
  real y[m];              // RAF
  real<lower=0> mu; // mutation rate
  real B; // effect size
  int N; // population size
}
parameters {
  real<lower=1e-20, upper=1e-3> W;                // selection
}
model {
  y ~ my_raf_lpdf(W, N, mu, B);       // posterior 
}

```


Finally we will fit the stan model and get the posterior distribution on W  
```{r, warning=FALSE}
data <- list(y = x_freqs, m = m, B = B, N = N, W=W, mu=mu)
fit <- sampling(simple_RAF, data = data)
```

Let's look at the results. 
```{r}
print(fit)
plot(fit)
traceplot(fit,pars="W" )
```

Right now on default settings, I am not doing a great job of infering W (whole order of magnitude off).  

## Infer W using evolutionary status  

So far we have only considered inferring W using the distribution of risk allele frequencies. Now we will introduce evolutionary information. The probability that the risk allele is derived is:  

$$P(D|x, N, W, \beta) = \frac{e^{4N\beta W} - e^{4N\beta x}}{-1 + e^{4 N \beta W}}$$  

Here D is an indicator variable that equals 0 if the risk allele is ancestral and 1 if the risk allele is derived. Now we want to simulate data that consisits of a set of pairs of risk allele frequecies and D (if the risk allele is derived). We will use the same inverse transform sampling function from above to draw risk allele frequencies according to $psi$ and then use the functions below to assign evolutionary status (risk derived or ancestral) to each allele by drawing a bernoulle RV with probability calculate from the above equation.    
```{r}
# Probability that an allele is derived 
derived_risk <- function(x, W, N, B) {
  one <- exp(4 * N * B * W) - exp(4 * N * B * W * x)
  two <- -1 + exp(4 * N * B * W)
  return(one/two)
}

# 0 = Ancestral 
# 1 = Derived 
assign_status <- function(m,W,N,B,x_freqs) {
  status <- rep(NA, m)
  for (i in 1:m) {
    freq <- x_freqs[i]
    p <- derived_risk(freq, W=W, N=N, B=B)
    ind <- rbinom(1,1, p)
    status[i] <- ind
  }
  return(status)
}
```

Now we will assign evolutionary status to our previously simulated set of RAFs:   
```{r}
status <- assign_status(m, W, N, B, x_freqs)
# Look at distribution of status data
table(status)
```

We now have our data and wish to do inference on the W used to generate the data. We can write the posterior distribution on W as:  

$$P(W|x, N, \beta, D) \propto P(D|W, x, N, \beta) P(W) $$  

The likelihood is the same as written above and the prior $P(W)$ we will consider uniform and equal to 1. Now that we have the posterior distribution we can sample from it using STAN.   

Here is the STAN program I wrote to sample from the posterior:  
```{stan output.var="simple_status"}
functions {
  real my_derived_risk_lpdf(real f, real W, int N, real mu, real B){ 
    real prob;
    prob = (exp(4 * N * B * W) - exp(4 * N * B * W * f)) / (-1 + exp(4 * N * B * W)) ;
    return prob;
  }
  real my_ll_risk_lpmf(int[] s, real[] x, real W, int N, real mu, real B){
    vector[num_elements(x)] prob;
    real lprob;
    for (i in 1:num_elements(x)) {
      if (x[i] == 0) {
        prob[i] = 1 - my_derived_risk_lpdf(x[i] | W, N, mu, B);
      }
      else {
        prob[i] = my_derived_risk_lpdf(x[i] | W, N, mu, B);
      }
    }
    lprob = sum(log(prob));
    return lprob;
  }
}
data {
  int<lower=0> m;         // Number snps
  real x[m];              // RAF
  int s[m];
  real<lower=0> mu; // mutation rate
  real B; // effect size
  int N; // population size
}
parameters {
  real<lower=1e-20, upper=1e-3> W;                // selection
}
model {
  s ~ my_ll_risk_lpmf(x, W, N, mu, B);
}

```

Let's fit the STAN model to our data
```{r, warning=FALSE}
data <- list(s = status, x = x_freqs, m = m, B = B, N = N, W=W, mu=mu)
fit <- sampling(simple_status, data = data)
```

Let's look at the results 
```{r}
print(fit)
plot(fit)
traceplot(fit,pars="W" )
```

Again I am not doing a very good job. Using the evolutionary information consistently over estimates W.  

## Infer W using both RAF and Evolutionary Status  

So far we have been using only the RAF distribution or evolutionary status to compute the log-likelihood used in the inference. By factoring the likelihood, we are able to use both peices of information to calculate the posterior:  

$$P(W|x, D, N, \mu, \beta) \propto P(D|x, N, \beta, W) P(x|N, \beta, \mu) P(D) P(x)$$  

Here is the STAN program I wrote to sample from this posterior distribution. 
```{stan output.var="RAF_status"}
functions {
  real my_raf_lpdf(real f, real W, int N, real mu, real B){ // PSI density 
    real prob;
    prob = ((f * (1- f))^((4*N*mu) - 1)) * exp(-4 * N * B * W * f) ;
    return prob;
  }
  real my_derived_risk_lpdf(real f, real W, int N, real mu, real B){ // Probability density of derived risk
    real prob;
    prob = (exp(4 * N * B * W) - exp(4 * N * B * W * f)) / (-1 + exp(4 * N * B * W)) ;
    return prob;
  }
  real my_ll_risk_lpmf(int[] s, real[] x, real W, int N, real mu, real B){ // General log-likelihood 
    vector[num_elements(x)] prob;
    real lprob;
    for (i in 1:num_elements(x)) {
      if (x[i] == 0) {
        prob[i] = (1 - my_derived_risk_lpdf(x[i] | W, N, mu, B)) * my_raf_lpdf(x[i] | W, N, mu, B);
      }
      else {
        prob[i] = my_derived_risk_lpdf(x[i] | W, N, mu, B) * my_raf_lpdf(x[i] | W, N, mu, B);
      }
    }
    lprob = sum(log(prob));
    return lprob;
  }
}
data {
  int<lower=0> m;         // Number snps
  real x[m];              // RAF
  int s[m];          // A/D status (coded as 0 or 1)
  real<lower=0> mu; // mutation rate
  real B; // effect size
  int N; // population size
}
parameters {
  real<lower=1e-20, upper=1e-3> W;                // selection
}
model {
  s ~ my_ll_risk_lpmf(x, W, N, mu, B);
}

```

We can use the same data simulated above and fit our STAN model: 
```{r, warning=FALSE}
data <- list(s = status, x = x_freqs, m = m, B = B, N = N, W=W, mu=mu)
fit <- sampling(RAF_status, data = data)
```

Let's look at our results 
```{r}
print(fit)
plot(fit)
traceplot(fit,pars="W" )
```
