---
title: "PRF_Inference"
author: "Jennifer Blanc"
date: "1/9/2020"
output: html_document
---

```{r setup, include=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = TRUE)
rm(list = ls())
library(rstan)
options(mc.cores = parallel::detectCores())
rstan_options(auto_write = TRUE)
set.seed(14)
```

## Introduction  

For this project, we are interested in inferring the degree of mutational bias across traits. In order to calculate the mutational bias, we need to infer specific parameters from the data. First consider the polygenic mutation-selection balance model. In this model, the selection coefficient for each individual allele is determined by the approximation below:  

$$s \approx \beta \phi(T) S$$  

Here $\beta$ is the true effect size of the allele, $\phi(T)$ is the height of the liability threshold, and S is the fitness cost of the disease. We can think of $\beta * \phi(T)$ as an approximation of the portion of the population pushed over the liability threshold when the protective allele is swapped to the risk allele. $\phi(T) S$ is a compound parameter that is related to the prevalence of the disease. The larger the $\phi(T)$ the higher the prevalence of disease (the threshold is closer to the center), the smaller the fitness cost must be. Here we will call this compound parameter $\eta$:  

$$\eta = \phi(T)S$$  

In this analysis we are interested in inferring from simulated data. We will simulate data under a specific value of $\eta$ and then use both maximum likelihood and Bayesian inference (using STAN) to infer back the value of $\eta$, the data were generated under.   

## Infer eta using RAF   

To start with we are going to make a (very) simplifying assumption that we know all the causal loci and that they all have an effect size of one ($\beta$ = 1). W will use the poison random field model to generate the data. Under this model the probability of observing a site at a certain risk allele freq, x, is given by:  

$$\Psi(x |\eta, N, \beta, \mu) = \frac{4N\mu}{x(1-x)}\frac{e^{-4 N\beta \eta x}}{1 + e^{-4N\beta \eta}}$$

Let's plot this function for difference values of $\eta$:  
```{r}
psi <- function(x, eta, B, N, mu) {
  top <- exp(-4*N*eta*B*x) * (4*N*mu)
  bottom <- (1 + exp(-4*N*eta*B)) * x * (1-x)
  return(top/bottom)
}

par(mfrow=c(2,2))
curve(psi(x,1e-6,1,10000,1e-8), col = "red", xlim = c(0,1), ylab = "", main = "4Ns = 0.04")
curve(psi(x,1e-5,1,10000,1e-8), col = "blue", xlim = c(0,1), ylab = "", main = "4Ns = 0.4")
curve(psi(x,1e-4,1,10000,1e-8), col = "green", xlim = c(0,1), ylab = "", main = "4Ns = 4")
curve(psi(x,1e-4,1,10000,1e-8), col = "purple", xlim = c(0,1), ylab = "", main = "4Ns = 40")
```

We can see that for larger values of $4Ns$ (where s = $\beta \eta$), the risk allele is held at lower frequencies. This intuitively makes sense and confirms that we have our signs correct in our model.  

Now we want to simulate data under a specific value of $\eta$ using inverse transform sampling.  
```{r}
simulate_data <- function(N, B, eta, mu, m) {
  # This function will sample a set of risk allele frequencies 
  
  # Get probability that a site is segregating 
  denominator <- integrate(psi, (1/(4*N)), (1- (1/(4*N))), eta=eta, B=B, N=N, mu=mu)$value 
  
  # Discretize the target CDF 
  i_values <- seq(1/(4*N),1-(1/(4*N)), 1/(2*N))
  pis <- rep(0, length(i_values))

  # Integrate up to i + (0.5/2N) for all i 
  for (j in 1:(length(i_values)-1)) {
    pis[j] <- integrate(psi, i_values[1], i_values[j+1], N=N, B=B, eta=eta, mu=mu)$value / denominator
  }
  
  # Draw from uniform
  U <- runif(m)

  # Find smallest i such that pi < U is true
  UT <- rep(0, length(U))
  for (j in 1:length(U)) {
    UT[j] <- which(U[j] <= pis)[1] 
  }
  
  # Calculate set of RAF 
  x_freqs <- UT / (2*N)
  
  return(x_freqs)
}
```

Now we are ready to simulate data. We will set the constant parameters we are not interested in inferring and then set $\eta$ and simulate the data   
```{r}
# Constant parameters 
N <- 10000
mu <- 10^-8 
B <- 1
m <- 500 # Number of causal snps 

# Value of eta for simuation 
eta <- 1e-5

# Simulate data 
x_freqs <- simulate_data(N, B, eta, mu, m)

# Look at distribution of data compared to true distribution 
par(mfrow = c(1,2))
hist(x_freqs)
curve(psi(x,eta=eta,N=N,B=B, mu=m), col = "red", xlim = c(0,1), ylab = "")
```


Now we have data are are ready to do inference. Let's look again at the distribution of risk alleles as stated above: 

$$\Psi(x | \eta,N,\beta, \mu) = \frac{4N\mu}{x(1-x)}\frac{e^{-4 N\beta \eta x}}{1 + e^{-4N\beta \eta}}$$  
We want to write the likelihood of observing our data given a set of parameters. We can do this using by integrating of probability of observing the observed allele frequency for each site and multiplying across sites:      

$$L = \prod\limits_{l=1}^L P(X_i = x,| \eta, N, \beta, \mu) = \prod\limits_{l=1}^L \int\limits_{x_i-\frac{1}{4N}}^{x_i + \frac{1}{4N}} \frac{4N\mu}{x_i(1-x_i)}\frac{e^{-4 N\beta \eta x_i}}{1 + e^{-4N\beta \eta}}$$  

Below is a function to calculate the log likelihood:  
```{r}
log_like <- function(eta, data, N, B, mu, m) {
  snps <- rep(NA, m)
  for (i in 1:m) {
    freq <- data[i]
    snps[i] <- log(integrate(psi, freq-(1/(4*N)), freq+(1/(4*N)), eta=eta, N=N, B=B, mu=mu)$value)
  }
  return(sum(snps))
}
log_like(data=x_freqs, N=N,B=B,mu=mu,eta=eta, m=m)
```

Now I will try to calculate the maximum likelihood estimator in a super simple way - I'll just try feeding in a range of eta values and seeing which one maximizes the log-likelihood.  
```{r}
# Start with a large range of eta values 
eta_range <- seq(0, 2e-5, by=1e-7)

# Loop through and calculate the LL for each eta value 
result_LL <- rep(NA, length(eta_range))
for (i in 1:length(eta_range)) {
  eta_test <- eta_range[i]
  result_LL[i] <- log_like(data=x_freqs, N=N, B=B, mu=mu, m=m, eta=eta_test)
}

plot(y=result_LL, x = eta_range)
abline(v = eta_range[which.max(result_LL)], col = "red")
abline(v = eta, col = "green")

print(paste0("MLE eta ", eta_range[which.max(result_LL)]))
print(paste0("True eta ", eta))
```

Above we got a maximum likelihood estimation for $\eta$ using a simple grid search. Now we will try to get an estimate of $\eta$ using a Bayesian approach using STAN. Let's look again at the distribution of risk alleles as stated above:  

$$\Psi(x | \eta, N, \beta, \mu) = \frac{4N\mu}{x(1-x)}\frac{e^{-4 N\beta \eta x}}{1 + e^{-4N\beta \eta}}$$ 

Using Bayes rule we can re-write this as:  

$$P(\eta | x, N, \beta, \mu) = \frac{P(x| \eta, N, \beta, \mu) P(\eta)}{P(x|N, \beta, \mu)} \propto P(x| \eta, N, \beta, \mu) P(\eta)$$  

Now we have a posterior distribution on eta that we can sample from using the likelihood ($\Psi(x | \eta, N, \beta, \mu)$) and the prior, which we for now assume to be 1. Below is the STAN program I wrote to sample from this posterior    

```{stan, output.var="simple_RAF"}
functions {
  real my_raf_lpdf(real[] x, real eta, int N, real mu, real B){ // PSI density 
    vector[num_elements(x)] prob;
    real lprob;
    for (i in 1:num_elements(x)) {
      prob[i] = ((4*N*mu)/ (x[i]*(1-x[i]))) * (exp(-4*N*B*eta*x[i])/(1+exp(-4*N*B*eta)));
    }
    lprob = sum(log(prob));
    return lprob;
  }
}
data {
  int<lower=0> m;         // Number snps
  real y[m];              // RAF
  real<lower=0> mu; // mutation rate
  real B; // effect size
  int N; // population size
}
parameters {
  real<lower=1e-20, upper=1e-3> eta;                // selection
}
model {
  y ~ my_raf_lpdf(eta, N, mu, B);       // posterior 
}

```

Finally we will fit the stan model and get the posterior distribution on eta  
```{r, warning=FALSE, result='hide'}
data <- list(y = x_freqs, m = m, B = B, N = N, eta=eta, mu=mu)
fit <- sampling(simple_RAF, data = data)
```

Let's look at the results. 
```{r}
print(fit)
plot(fit)
traceplot(fit,pars="eta" )
```

We can see that the estimate of $/eta$ is pretty close to the true value of 1e-5.  

## Infer eta using evolutionary status  

So far we have only considered inferring $\eta$ using the distribution of risk allele frequencies. Now we will introduce evolutionary information. The probability that the risk allele is derived is:  

$$P(D|x, N, \eta, \beta) = \frac{e^{4N\beta \eta} - e^{4N\beta x}}{-1 + e^{4 N \beta \eta}}$$  

Here D is an indicator variable that equals 0 if the risk allele is ancestral and 1 if the risk allele is derived. Now we want to simulate data that consists of a set of pairs of risk allele frequencies and D (if the risk allele is derived). We will use the same inverse transform sampling function from above to draw risk allele frequencies according to $\Psi$ and then use the functions below to assign evolutionary status (risk derived or ancestral) to each allele by drawing a Bernoulli RV with probability calculate from the above equation.  

```{r}
# Probability that an allele is derived 
derived_risk <- function(x, eta, N, B) {
  one <- exp(4 * N * B * eta) - exp(4 * N * B * eta * x)
  two <- -1 + exp(4 * N * B * eta)
  return(one/two)
}

# 0 = Ancestral 
# 1 = Derived 
assign_status <- function(m,eta,N,B,x_freqs) {
  status <- rep(NA, m)
  for (i in 1:m) {
    freq <- x_freqs[i]
    p <- derived_risk(freq, eta=eta, N=N, B=B)
    ind <- rbinom(1,1, p)
    status[i] <- ind
  }
  return(status)
}
```

Now we will assign evolutionary status to our previously simulated set of RAFs:   
```{r}
status <- assign_status(m, eta, N, B, x_freqs)
# Look at distribution of status data
table(status)
```

Now our data is comprised of RAF stored in "x_freqs" and whether the risk allele is derived stored in "status". We are ready to compute the likelihood:  

$$L = \prod\limits_{i=1}^M \frac{e^{4N\beta eta}-e^{4N\beta eta x_i}}{-1 + e^{4N\beta eta}}$$  

```{r}
log_like_status <- function(eta, data_freq, data_status, N, B, m) {
  snps <- rep(NA, m)
  for (i in 1:m) {
    freq <- data_freq[i]
    stat <- data_status[i]
    if (stat == 1) {
      snps[i] <- log(derived_risk(x=freq, eta=eta, N=N, B=B))
    } else {
      snps[i] <- log(1 - derived_risk(x=freq, eta=eta, N=N, B=B))      
    }
  }
  return(sum(snps))
}
log_like_status(data_freq=x_freqs,data_status=status,N=N,B=B,eta=eta,m=m)
```

Now let's use a grid search to find the MLE of $\eta$:  
```{r}
# Start with a large range of eta values 
eta_range <- seq(0, 3e-5, by=1e-7)

# Loop through and calculate the LL for each eta value 
result_LL <- rep(NA, length(eta_range))
for (i in 1:length(eta_range)) {
  eta_test <- eta_range[i]
  result_LL[i] <- log_like_status(data_freq=x_freqs, data_status = status, N=N, B=B, m=m, eta=eta_test)
}

plot(y=result_LL, x = eta_range)
abline(v = eta_range[which.max(result_LL)], col = "red")
abline(v = eta, col = "green")

print(paste0("MLE eta ", eta_range[which.max(result_LL)]))
print(paste0("True eta ", eta))
```


We now have an MLE estimate of $\eta$. Similarly to what we did for risk allele frequency, we want to try to do Bayesian inference using STAN. We can write the posterior distribution on eta as:  

$$P(\eta|x, N, \beta, D) \propto P(D|\eta, x, N, \beta) P(\eta) $$  

The likelihood is the same as written above and the prior $P(\eta)$ we will consider uniform and equal to 1. Now that we have the posterior distribution we can sample from it using STAN.   

Here is the STAN program I wrote to sample from the posterior:  
```{stan output.var="simple_status"}
functions {
  real my_derived_risk(real f, real eta, int N, real mu, real B){ 
    real prob;
    prob = (exp(4 * N * B * eta) - exp(4 * N * B * eta * f)) / (-1 + exp(4 * N * B * eta)) ;
    return prob;
  }
  real my_ll_risk_lpmf(int[] s, real[] x, real eta, int N, real mu, real B){
    vector[num_elements(x)] prob;
    real lprob;
    for (i in 1:num_elements(x)) {
      if (s[i] == 0) {
        prob[i] = 1 - my_derived_risk(x[i], eta, N, mu, B);
      }
      else {
        prob[i] = my_derived_risk(x[i], eta, N, mu, B);
      }
    }
    lprob = sum(log(prob));
    return lprob;
  }
}
data {
  int<lower=0> m;         // Number snps
  real x[m];              // RAF
  int s[m];
  real<lower=0> mu; // mutation rate
  real B; // effect size
  int N; // population size
}
parameters {
  real<lower=1e-20, upper=1e-3> eta;                // selection
}
model {
  s ~ my_ll_risk_lpmf(x, eta, N, mu, B);
}

```

Let's fit the STAN model to our data
```{r, warning=FALSE, results='hide'}
data <- list(s = status, x = x_freqs, m = m, B = B, N = N, eta=eta, mu=mu)
fit <- sampling(simple_status, data = data)
```

Let's look at the results 
```{r}
print(fit)
plot(fit)
traceplot(fit,pars="eta" )

```

We are doing a decent job of inferring $/eta$ using the evolutionary status of the risk allele.  

## Infer eta using both RAF and evolutionary status  

Finally we can combine both pieces of information to infer $\eta$. To do this we simply need to factorize the likelihood:  

$$L = \prod\limits_{l=1}^M P(x_i|\eta, N, \beta, \mu) P(D_i |x_i,\eta, N, \beta)$$  

Let's use this likelihood to do maximum likelihood estimation:  
```{r}
# Here we will calculate the log likelihood to avoid getting too small numbers 
log_like_combined <- function(eta, data_freq, data_status, N, B,mu, m) {
  snps <- rep(NA, m)
  for (i in 1:m) {
    freq <- data_freq[i]
    stat <- data_status[i]
    if (stat == 1) {
      snps[i] <- log(derived_risk(x=freq, eta=eta, N=N, B=B))
    } else {
      snps[i] <- log(1 - derived_risk(x=freq, eta=eta, N=N, B=B))      
    }
    snps[i] <- snps[i] + log(integrate(psi, freq-(1/(4*N)), freq+(1/(4*N)), eta=eta, N=N, B=B, mu=mu)$value) # RAF Info 
  }
  return(sum(snps))
}
log_like_combined(data_freq=x_freqs,data_status=status,N=N,B=B,eta=eta,m=m, mu=mu)
```

Now let's use a grid search to find the MLE of $\eta$:  
```{r}
# Start with a large range of eta values 
eta_range <- seq(0, 2e-5, by=1e-7)

# Loop through and calculate the LL for each eta value 
result_LL <- rep(NA, length(eta_range))
for (i in 1:length(eta_range)) {
  eta_test <- eta_range[i]
  result_LL[i] <- log_like_combined(data_freq=x_freqs, data_status = status, N=N, B=B, m=m, eta=eta_test, mu=mu)
}

plot(y=result_LL, x = eta_range)
abline(v = eta_range[which.max(result_LL)], col = "red")
abline(v = eta, col = "green")

print(paste0("MLE eta ", eta_range[which.max(result_LL)]))
print(paste0("True eta ", eta))
```
 

Finally, we will again use STAN to sample from the factorized posterior:  

$$P(X,D|\eta,N, B, \mu) \propto P(X|\eta,N, B,\mu)P(D|\eta, N, B)$$   

```{stan output.var="simple_combined"}
functions {
  real my_derived_risk(real f, real eta, int N, real mu, real B){ 
    real prob;
    prob = (exp(4 * N * B * eta) - exp(4 * N * B * eta * f)) / (-1 + exp(4 * N * B * eta)) ;
    return prob;
  }
  real my_ll_risk_lpmf(int[] s, real[] x, real eta, int N, real mu, real B){
    vector[num_elements(x)] prob;
    real lprob;
    for (i in 1:num_elements(x)) {
      if (s[i] == 0) {
        prob[i] = 1 - my_derived_risk(x[i], eta, N, mu, B);
      }
      else {
        prob[i] = my_derived_risk(x[i], eta, N, mu, B);
      }
    }
    lprob = sum(log(prob));
    return lprob;
  }
  real my_raf_lpdf(real[] x, real eta, int N, real mu, real B){ // PSI density 
    vector[num_elements(x)] prob;
    real lprob;
    for (i in 1:num_elements(x)) {
      prob[i] = ((4*N*mu)/ (x[i]*(1-x[i]))) * (exp(-4*N*B*eta*x[i])/(1+exp(-4*N*B*eta)));
    }
    lprob = sum(log(prob));
    return lprob;
  }
}
data {
  int<lower=0> m;         // Number snps
  real x[m];              // RAF
  int s[m];
  real<lower=0> mu; // mutation rate
  real B; // effect size
  int N; // population size
}
parameters {
  real<lower=1e-20, upper=1e-3> eta;                // selection
}
model {
  s ~ my_ll_risk_lpmf(x, eta, N, mu, B);
  x ~ my_raf_lpdf(eta, N, mu, B); 
}

```

Let's fit the STAN model to our data
```{r, warning=FALSE, results='hide'}
data <- list(s = status, x = x_freqs, m = m, B = B, N = N, eta=eta, mu=mu)
fit <- sampling(simple_combined, data = data)
```

Let's look at the results 
```{r}
print(fit)
plot(fit)
traceplot(fit,pars="eta" )

```








